A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2020-04-03 paper/paper-revision.Rnw
Add scanner images to paper (as figure) and small text edits

In the United States, forensic analysis of shoe print evidence in criminal cases is varied and difficult. The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis (i.e., one based on randomly acquired characteristics) when a print is partial or of low quality. In fact, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. %\footnote{Leslie Hammer, presentation to CSAFE on March 5, 2018}).
In cases where individualizing characteristics are not used, it still may be both possible and useful to assess the random match probability of the footwear evidence at the level of class characteristic; in other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern? Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns (e.g., knockoffs, similar models over different years, etc).

From a statistical perspective, random match probability could theoretically be calculated by sampling a large number of shoes from a given population and calculating the proportion of shoes in the sample whose tread patterns are similar to the given print. In practice, one potential method of sampling could be a low-profile scanner placed in a high traffic area. Such a scanner could take \mt{impressions or} pictures of shoes walking past, and those scans could then be used to create a database of shoes relevant to the local area.

However, studies by \citet{parent_sandy_significance_nodate} and \citet{benedict_geographical_2014} demonstrate that even \fix{[a couple hundred]} shoe prints constitute an unweildy amount of data requiring \fix{[a large amount of data processing]}. Thus, in parallel with the development of such a scanner, it is necessary to develop a way to process large amounts of data automatically. \mt{The primary focus of this paper is the development of one such method for large-scale automatic data processing, where "data processing" refers to assessing} the similarity of the shoe tread patterns in the collected data.

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, early prototypes of a footwear scanner have suggested that cameras are more robust than pressure scanners for placement in public locations that may experience inclement weather or other non-controlled conditions. \mt{Thus, we are working to develop a scanner using photo technology} \fix{(reference/cite image of the scanner)}. \mt{This scanner would take pictures of shoe outsoles, much like images that are currently found on the websites of online shoe retailers \fix{(reference/cite test image of  an outsole on the scanner)}. Although the images from the scanner are more degraded in some ways than those found from online retailers, the fundamental requirements to analyze such data would be similar.} Therefore, \fix{we move forward by considering how to assess similarity of tread patterns in images of outsoles, in particular.} More specifically, this assessment is done by identifying a set of relevant features and using image analysis methods to detect the presence or absence of the features in each image, and then defining similarity measures based on those features to determine how many shoes have a tread pattern similar to the pattern of interest.

>>>>

In the United States, forensic analysis of shoe print evidence in criminal cases is varied and difficult. The process of collecting print evidence from a crime scene is not standardized, and it is often impossible to conduct an individualizing analysis (i.e., one based on randomly acquired characteristics) when a print is partial or of low quality. In fact, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. In the cases where individualizing characteristics are not used, it may be both possible and useful to assess the random match probability of the footwear evidence at the level of class characteristics; in other words, given a questioned shoe print, what is the probability that a randomly selected shoe in the relevant population has a similar tread pattern? Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns (e.g., knockoffs, similar models over different years, etc).

From a statistical perspective, random match probability could theoretically be calculated by sampling a large number of shoes from a given population and calculating the observed proportion of shoes in the sample whose tread patterns are similar to the given print. In practice, one potential method of sampling could be a low-profile scanner placed in a high traffic area. Such a scanner could take \mt{impressions or} pictures of shoes walking past, and those scans could then be used to create a database of shoes relevant to the local area.

However, studies by \citep{parent_sandy_significance_nodate} and \citep{benedict_geographical_2014} demonstrate that even a couple hundred shoe prints constitute an unweildy amount of data that then requires \fix{[a large amount of data processing]}. Thus, in parallel with the development of such a scanner, it is necessary to develop a way to process large amounts of data automatically. \mt{The primary focus of this paper is the development of one such method for large-scale automatic data processing, where "data processing" refers to assessing} the similarity of the shoe tread patterns in the collected data.

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, early prototypes of a footwear scanner have suggested that cameras are more robust than pressure scanners for placement in public locations that may experience inclement weather or other non-controlled conditions. \mt{Thus, we are working to develop a scanner using photo technology. This scanner would take pictures of shoe outsoles, much like images that are currently found on the websites of online shoe retailers \autoref{fig:prototype-scanner}. Although the images from the scanner are more degraded in some ways than those found from online retailers, the fundamental requirements to analyze such data would be similar.} Therefore, \fix{we move forward by considering how to assess similarity of tread patterns in images of outsoles, in particular.} More specifically, this assessment is done by identifying a set of relevant features and using image analysis methods to detect the presence or absence of the features in each image, and then defining similarity measures based on those features to determine how many shoes have a tread pattern similar to the pattern of interest.

\fix{How to make the scanner images a figure instead of a table?}

\begin{figure}
\centering
\setlength\tabcolsep{1mm}
\begin{tabular}{cc}
\raisebox{-.5\height}{\includegraphics[width=.45\textwidth]{images/prototype/20200402_151639.jpg}} &
\raisebox{-.5\height}{\includegraphics[width=.45\textwidth]{images/prototype/GOPR1231-blur-enhance_square.jpg}} \vspace{1mm}\\

\end{tabular}
\caption[Images of prototype scanner.]{Images of the prototype scanner for collecting a large number of outsole images.}\label{fig:prototype-scanner}
\end{figure}
